### 2011-10-13 doc/manuscript.tex
Functional categories are being discussed in the manuscript.

Bacteria evolves new biological features through horizontally transferred genes
from other lineages and vertically transmitted mutations on genes within the
same lineage.  While mutations on a gene modifiy relatively small portions of the 
gene through substitutions, insertions, and deletions of deoxyribonucleic acids,
gene transfers can reshape the organization of a bacterial genome,
which could revamp biological functions of the species.  Recipient bacteria of
transferred genes can substitute transferred genes for homologous parts of their
genomes, or integrate transferred genes into their genomes without replacement.
These two processes are here in this manuscript discriminated by calling the
former \textit{recombining gene transfer}, and the latter \textit{horizontal
gene transfer} \citep{Ochman2001,Lawrence2009}.  Despite the two distinct
evolutionary processes in bacteria, analyses of complete genome sequences in
efforts of better understanding bacterial evolution have paid less attention to
a combined approach to illuminating disparity of the two gene transfer
processes. We attempted here to use and compare two approaches for
distinguishing recombining gene transfer process from horizontal, or
non-recombining, gene transfer, and tackled the two processes in association with
virulence genes.

Homologous recombining gene transfer could be studied more readily using the
core genomes of individual bacteria while horizontal gene transfer might be more
easily detected using the pan genomes. Recently, a method of inferring
homologous recombination with whole bacterial genomes was developed
\citep{Didelot2010}. The model of homologous recombination allowed them to study
flows of genetic segments between bacteria individuals of \textit{Bacillus
cereus} species in the core genomes.  Other evolutionary processes acting on
bacterial genomes include genome rearragement, translocation, inversion,
duplication, gain, loss, etc.  These processes did not readily lend themselves
to statistically more sound model-based approaches such as maximum likelihood.
Instead, a parsimony-based approach to inferring gene duplication, loss, and
transfer events was developed by \citet{Doyon2011}.  We noted that the first
approach could be used to infer transfer of genomic segments due to homologous
recombination, and the second could help similarly attributed to horizontal gene
transfer.

In short, the complete genomes of the two spcies of SPY and SDE presented us a
unique opportunity to study gene transfer over relatively recent time scales.
We took advantage of the model-based approach for inferring homologous
recombination, and the parsimony-based one for detecting gene duplication, loss,
and horizontal transfer. We compared both of the results in recombining gene
transfer events to show that the parimony-based method inferred recombining gene
transfer events in biological functional categories similar to those inferred
with the model-based approach. This was the first attempt as far as we know to
analyze SPY and SDE in genome-wide scale for detecting gene transfer events not
only in the core genome but also in the pan genome, which made the analysis
complete in terms of the amount of used genome data. Our analysis of core
genomes also showed for the first time that the net directionality of
recombining gene transfer from SPY to SDE was stronger than the reverse
direction.

@@

Five genomes except \textit{S.\ equi} ssp.\textit{\ equi} strain 4047 were
chosen (Table S\ref{tab:genome}) to study recombining gene transfer using the
model-based approach.  A genome alignment of the five genomes showed two
distinct genome structures of SPY and SDE (Figure S\ref{fig:mauve}). The SDE and
SDD genomes were larger in length than those of SPY. Although it was unclear
whether SDD was different from either of SDE1 and SDE2 from the graphical view
of the genome alignment, a spcies tree inferred from the genome alignment
showed that SDD branch was attached to the clade branch of SDE1 and SDE2 (Figure
S\ref{fig:tree5}). The distribution of the alignment block lengths was skewed to
the right with mean of 4188 base pairs long (Figure S\ref{fig:blocksize}), which
supported that the model assumption of geometric block length distribution might
be plausible in \texttt{ClonalOrigin} method (see section \textbf{Materials and
Methods} and \citealt{Didelot2010}).  The relative strength ($r/m$) of
recombination compared to mutations indicated that recombination-driven
substitutions affected more sites than did mutation-driven substitutions (Table
S\ref{tab:clonalframe}). This could imply that in the \textit{Streptococcus}
species bacterial recombination process appeared to accelerate molecular
evolution of more sites than did point mutations process.

\subsubsection{Recombination hotspots were observed in genomes}

Along the genome of SDE1 population mutation rate did not appear to change much
(Figure S\ref{fig:scatter3}).  We found a few spots of genomic regions where the
mutation rates were larger than twice the global estimate of mutation rate.  The
recombinant tract length changed in a way that we could not note any particular
positions with notable tract length values compared with other positions along
the genome.  As recombinant tract lengths were smaller, recombination rates per
site seemed larger (Figure S\ref{fig:rhologdelta}).  The recombination rate
appeared to change more dynamically. Especially, a genomic region of SDE1 near
200 kilobase pairs from the origin of replication contained alignment blocks with
bell-shaped recombination rates.

\subsection{SPY and SDE were more closely related in about tenth of the genome}

There are 105 possible rooted topologies relating five taxa.  We expected the
dominant topology across sampled genealogies to be concordant with that of the
species tree.  Indeed, two thirds of the local gene tree topologies (see section
\textbf{Materials and Methods}) along the alignment blocks were the same as the
species tree, which allowed us to be more confident of the species tree inferred
using the alignment blocks (Table S\ref{tab:Gene-tree-topologies}).  Other
frequently sampled topologies were expected to be potentially informative about
common recombination events.  The second most frequent topology (9.3\%) was one
where SDD was the outgroup, and SPY and SDE were sister clades. Therefore, SPY
and SDE appeared be more closely related with each other than each to SDD's
genome in about tenth of the core genome. The other topologies had less than 4\%
of proportions.

@@

The heat map of Figure \ref{fig:Heatmap-of-recombination} (see Tables
S\ref{tab:heatmap} and S\ref{tab:obsheatmap} for the heat map values and the
observed number of recombinant edges) shows the log base 2 ratios of the
\textit{a posteriori} number of recombinant edges for a pair of donor and
recipient species tree branches and their corresponding expected values given
the species tree (Figure S\ref{fig:tree5}) and the three population parameters
(Table S\ref{tab:three}).  Because recombinant edges from any species tree
branch to the root branch are impossible (see Figure S\ref{fig:clonalorigin}),
the column for the root as recipient was colored green, indicating invalid
recombinant edges.  Similarly, all of the cells for which a recipient branch was
older than the donor, and the donor and recipient branches did not share any contemporary time were
colored green. The diagonal cells were colored gray, indicating no significant
number of recombinant edges, because the observed and expected numbers of
recombinant edges were similar. As recombinant edges starting from and ending in
the same species tree branch would not change the species tree topology, which
might not
change the likelihood much, the cells would be colored gray.  Because SDD is a
strictly veterinary pathogenic bacteria, we did not expect that SDD received
much gene transfer from either SPY1 or SPY2: the log base 2 ratio from SPY1 to
SDD was -2.20, and that from SPY2 to SDD was -1.97.  On the contrary, the
recombing gene transfer from SDE1 to SDD was 0.756, and that from SDE2 to SDD
was 0.698. This relatively higher recombining gene transfer from two SDE
individuals to SDD compared with that from two SPY individuals to SDD reflected
that the SDD was more closely related to SDE than to SPY in most of the genomes,
and the genomic organization between SDD and SDE1/SDE2 were much similar
(Figure S\ref{fig:mauve}).  However, the gene transfer from SDD to either SDE1
or SDE2 was less frequent than the case of reverse direction; SDD effectively
served as a recipient rather than a donor. It was not unclear why SDD received
more genes from SDE1 or SDE2 than did SDE1 or SDE2 from SDD.

Because it was expected that the recombining gene transfer within a clade would
be more frequent than that between clades, not surprisingly the recombining gene
transfer log base 2 ratios between SDE1 and SDE2 (1.17 and 1.20) were about
twice as large as that from SDE1 and SDE2 to SPY1 and SPY2 (0.67, 0.66, 0.60,
and 0.39).  The intensities of recombining gene transfer from external branches
of SDE1 and SDE2 to external branches of SPY1 and SPY2 was similar to those from
SDE1 or SDE2 to SDD (0.76 and 0.70).  After all, either SPY1 or SPY2, and SDD as
recipient appeared to be no different from the view point of SDE1 or SDE2 as
donor since the similar values of recombining gene transfer intensity from SDE1
or SDE2 to one of the other three external branches. Surprisingly, the recombining
gene transfer log base 2 ratios between SPY1 and SPY2 (1.06 and 1.07) was
comparable with, yet even less than those from SPY1 and SPY2 to SDE1 and SDE2
(1.25, 1.25, 1.17, and 1.29).  Effectively, the recombining gene transfer from a
clade SPY1 or SPY2 to a different clade SDE1 or SDE2 seemed to be no different
from that within the clade of SPY1 and SPY2.  The two external branches of SDE
received genes from either SPY1 or SPY2 via recombining gene transfer as if SDE1
and SDE2 were in the same clade as SPY1 and SPY2.  This was more surprising
because the bacterial recombination between SPY and SDE seemed to happen more
often than that between SDE and SDD even though the genome configurations of SDE
and SDD were more similar than each to SPY.  However, this result could be
concordant with the fact that SPY and SDE shared the same ecological niche
despite their dissimilar genomic organization.  The gene transfer log base 2
ratio from SPY1 and SPY2 to SDE branch was also larger than any type of
recombining gene transfer: 1.88 and 1.94, respectively.  Additionally we found
58 genes for which a recombinant edge of certain types were sampled with
probability at least 0.9 by focusing on specific types of recombinant edges
(Table S\ref{tab:genes-transfer}).  Many of the genes in the list were
transferred from SPY to SDE rather than the reverse direction overall.  All
things considered, this epitomized that the recombining gene transfer was biased
toward the net direction from SPY to SDE.

We tried in vain to quantify differential horizontal gene transfer between SPY
and SDE with the parsimony-based approach (Figure \ref{fig:hgt-heatmap}).  Among
the clade-crossing transfers, we did not see any significant enrichment for a
specific direction, as opposed to the result for recombining transfers from the
model-based analysis (Figure S\ref{fig:mowgli-recomb-heatmap}).  SDE1 and SDE2
shared the most events (26 and 34) compared to other sample pairs.  SDD appeared
to acquire many genes from the ancestral SDE branch.  Transfer events between
the SPY and SDE clades were inferred in smaller numbers (<6) compared to those
within each clade. We were puzzled by this finding because virulence
genes were found to be associated with horizontal gene transfer, and yet we could
not show net directionality of horizontal gene transfer between SPY and SDE.

\subsection{More frequent recombining gene transfer than horizontal gene transfer}

Overall, we inferred 304 duplications, 944 recombining gene transfers, and 294
horizontal gene transfers, comprising 23.7\% of total transfer events (Figure
\ref{fig:Gene-duplication-loss}).  Among the 2314 gene families the largest
family consisted of 59 genes, while nearly half of gene families (1066) had
exactly one gene per species, or six genes total (Figure S\ref{fig:famsizes}).
Within clades we found that duplications made up only a small fraction of the
events  (10.2\%), while horizontal transfers and losses were significantly more
frequent (41.7\% and 48.0\%, respectively). Duplications were most often
species-specific.  The species SDD and SDE were the most expanded, both in terms
of duplications and in-bound transfers.

@@

About 10\% of the genes comprised of virulence facters, about half of which were
found in the core genome of the five genome alignment. Using the UCSC genome
browser tracks we located virulence genes with notable posterior recombination
probability.  These genes were by no means comprehensive list of virulence
factors that might have been transferred via homologous recombination.
Stressful host conditions for pathogeneic bacteria can include heat shock,
starvation, and iron depletion.  Iron uptake of bacteria was known to be
essential for survival of the bacteria under hostile host condition. We found
that parts of putative calcium transporter (MgtA) in SDE1 could have been
transferred from SPY1 (SpyM3\_0440).  An extracellular protein, putative
dipeptidase (SpyM3\_0465) that could be important for the survival of bacteria
to scavenge dipeptides from the host, was inferred to be transferred from SPY to
SDE: the whole gene was transferred, which was a case for the finding that
virulence genes tended to be transferred as a whole unit \citep{Chan2009}.
Another example of whole gene transfer was superoxide dismutase that was
reported to be associated virulence of \textit{Streptococcus agalactiae}
\citep{Poyart2001}: the whole gene including its 5' UTR, (SpyM3\_1071), was
affected by recombining gene transfer from SPY to SDE.  Some of the genes were
only affected in their parts.  The middle parts of an adhesins gene, putative
internalin A precursor (SpyM3\_1035), was found to be affected by recombining
gene transfer from SPY to SDE.  The middle part of a chaperonin GroEL, putative
heat shock protein (SpyM3\_1765), was inferred to transfer from SPY to SDE1:
increase of GroEL expression was known to allow for the bacteria to survive the
hostile host environment of acid shock.  Parts of the gene (SpyM3\_1749),
pyruvate formate lyase that was recently reported to have virulence
\citep{Yesilkaya2009}, seemed to be affected by recombining gene transfer.
Under starvation condition bacteria must use more effective energy metabolism: a
putative glucose kinase, (SpyM3\_1180), were inferred to be transferred from SPY
to SDE to adapt SDE's energy metabolism to survive such harsh condition.  We
also found that an MLST locus, called DNA mismatch repair protein (MutS),
appeared to be affected by recombining gene transfer.

\subsection{Virulence genes via horizontal gene transfer}

@@

Roughly, half of the virulence genes were in the pan genome.  We did not attempt
to comprehensively list virulence genes that were inferred to be transferred
horizontally. We rather sort virulence genes by the number of horizontal gene
transfer events to list the first several genes with high ranks.  Streptococcal
C5a peptidase (\textit{scpB}, SDEG\_0933) was inferred to have six horizontal
gene transfer.  Putative transposase were found to have a relatively large
number of horizontal gene transfer events (SpyM3\_1016, SpyM3\_1053,
SpyM3\_1054). Other genes with relatively large number of horizontal gene
transfer events included putative response regulator (SpyM3\_0174), putative
maltose operon transcriptional repressor (SpyM3\_0982), putative histidine
kinase (SpyM3\_0171), and putative cell wall hydrolase (SpyM3\_0731,
SpyM3\_0922, SpyM3\_1208).  Virulence genes encoding putative nucleases with a
secretion signal peptide were also found to have had experienced horizontal gene
transfer: genome-encoded streptodornase (SDEG\_0541), and prophage-associated
deoxyribonuclease (SDEG\_1103).  Protein F2 like fibronectin-binding protein
(SpyM3\_0104), an adhesin and invasin, was inferred to horizontally
transferred.

\subsection{Comparison between model-based and parsimony-based approaches}

Both of the model-based and parimony-based methods used discrepancy between a
reference species tree and gene trees to detect recombining gene transfer. In
essence the two methods worked in relatively similar manner. Firstly, functional
categories that were associated with recombining gene transfer were similar in
both methods (Tables S\ref{tab:functional-all} and
S\ref{tab:go-events-recombining}).  Secondly, both of them did not show
association of virulence genes with recombining gene transfer. Lastly, we found
a positive correlation of 0.39 with P-value of $2.2\times10^{-16}$ between the
recombination intensity and the number of recombining gene transfer for the
genes (Figure S\ref{fig:cmpcomowgli}).  The correlation (0.03 with P-value of
0.18) disappeared in the comparison between recombination intensity and number
of horizontal gene transfer events (Figure S\ref{fig:cmpcomowglihgt}).

\subsection{UCSC genome browser for bacterial gene transfer study}

We created a web-resource (Figure S\ref{fig:ucsc}) for gene transfer available at
\url{http://strep-genome.bscb.cornell.edu}.  For each of the five individual
genomes as a reference genome several annotation tracks were displayed:
alignment blocks, recombination rate per site for each block, local gene tree
topologies, recombination intensity, locations of virulence genes, posterior
probability of recombining gene transfer.  
% To Matt: what could we show in the pan-genome? or from Mowgli's analysis?

@@

The process of gene transfer may or may not move whole genes from one species to
another \citep[e.g.,][]{Chan2009}. We observed that units of recombining gene
transfer could happen with genes (Table S\ref{tab:genes-transfer}). In some
cases, parts of a gene were inferred to have been transferred, for instance,
transfer from SPY1 to SDE1 was inferred for 29\% of the aspartyl-tRNA synthetase
gene.  In other cases, the inferred transfer units covered a set of as much as
eight adjacent genes.  Of the genes listed in Table S\ref{tab:genes-transfer}
there were 12 groups of genes that recombinant tracts covered their gene
boudaries.  Most of the recombinant tracts often covered two genes. There were
transfer fragments that covered eight genes together: loci from SpyM3\_1518 to
SpyM3\_1525.

\subsection{Transporter, transferase, and aminoacyl tRNA synthetases}

>>>>

Bacteria evolves new biological features through horizontally transferred genes
from other lineages and vertically transmitted mutations on genes within the
same lineage.  While mutations on a gene modifiy relatively small portions of the 
gene through substitutions, insertions, and deletions of deoxyribonucleic acids,
gene transfers can reshape the organization of a bacterial genome,
which could revamp biological functions of the species.  Recipient bacteria of
transferred genes can substitute transferred genes for homologous parts of their
genomes, or integrate transferred genes into their genomes without replacement.
These two processes are here in this manuscript discriminated by calling the
former \textit{recombining gene transfer}, and the latter \textit{horizontal
gene transfer} \citep{Ochman2001,Lawrence2009}.  Despite the two distinct
evolutionary processes in bacteria, analyses of complete genome sequences in
efforts of better understanding bacterial evolution have paid less attention to
a combined approach to illuminating disparity of the two gene transfer
processes. We attempted here to use and compare two approaches for
distinguishing the recombining gene transfer process from the horizontal, or
non-recombining, gene transfer, and tackled the two processes in association with
virulence genes.

Homologous recombining gene transfer could be studied more readily using the
core genomes of individual bacteria while horizontal gene transfer might be more
easily detected using the pan genomes. Recently, a method of inferring
homologous recombination with whole bacterial genomes was developed
\citep{Didelot2010}. The model of homologous recombination allowed them to study
flows of genetic segments between bacteria individuals of \textit{Bacillus
cereus} species in the core genomes.  Other evolutionary processes acting on
bacterial genomes include genome rearragement, translocation, inversion,
duplication, gain, loss, etc.  These processes did not readily lend themselves
to statistically more sound model-based approaches such as maximum likelihood.
Instead, a parsimony-based approach to inferring gene duplication, loss, and
transfer events was developed by \citet{Doyon2011}.  We noted that the first
approach could be used to infer transfer of genomic segments due to homologous
recombination, and the second could be useful in studying horizontal gene
transfer.

In short, the complete genomes of the two spcies of SPY and SDE presented us a
unique opportunity to study gene transfer over relatively recent time scales.
We took advantage of the model-based approach for inferring homologous
recombination, and the parsimony-based one for detecting gene duplication, loss,
and horizontal transfer. Virulence genes were found to be more associated
non-recombining horizontal gene transfer than non-virulence genes.  We compared
both of the results in recombining gene transfer events to show that the
parimony-based method inferred recombining gene transfer events in biological
functional categories similar to those inferred with the model-based approach.
This was the first attempt as far as we know of systematically analyzing SPY and
SDE in genome-wide scale to detect gene transfer events not only in the core
genome but also in the pan genome, which made the analysis complete in terms of
the amount of used genome data. Our analysis of core genomes also showed for the
first time that the net directionality of recombining gene transfer from SPY to
SDE was stronger than the reverse direction.

@@

Five genomes except \textit{S.\ equi} ssp.\textit{\ equi} strain 4047 were
chosen (Table S\ref{tab:genome}) to study recombining gene transfer using the
model-based approach.  A genome alignment of the five genomes showed two
distinct genome structures of SPY and SDE (Figure S\ref{fig:mauve}). The SDE and
SDD genomes were larger in length than those of SPY. Although the graphical view
of the genome alignment did not clearly show whether SDD was different from
either of SDE1 and SDE2 in the genome organization, a spcies tree inferred from
the genome alignment showed that the external SDD branch was attached to the
internal clade branch of SDE1 and SDE2 (Figure S\ref{fig:tree5}). The
distribution of the alignment block lengths was skewed to the right with mean of
4188 base pairs long (Figure S\ref{fig:blocksize}), which supported that the
model assumption of geometric block length distribution might be plausible in
\texttt{ClonalOrigin} method (see section \textbf{Materials and Methods} and
\citealt{Didelot2010}).  The relative strength ($r/m$) of recombination compared
to mutations indicated that recombination-driven substitutions affected more
sites of the core genomes than did mutation-driven substitutions (Table
S\ref{tab:clonalframe}). This could imply that in the \textit{Streptococcus}
species bacterial recombination process appeared to accelerate molecular
evolution of more sites than did point mutations process.

\subsection{Recombination hotspots were observed in genomes}

Population mutation rate did not appear to change much across the genome of SDE1 
(Figure S\ref{fig:scatter3}).  We found a few spots of genomic regions where the
mutation rates were larger than twice the global estimate of mutation rate.  The
recombinant tract length changed in a way that we could not note any particular
positions with notable tract length values compared with other positions along
the genome.  As recombinant tract lengths were smaller, recombination rates per
site seemed larger (Figure S\ref{fig:rhologdelta}).  The recombination rate
appeared to change more dynamically. Especially, a genomic region of SDE1 near
200 kilobase pairs from the origin of replication contained alignment blocks with
bell-shaped recombination rates.

\subsection{SPY and SDE were more closely related in about tenth of the core genome}

There are 105 possible rooted labeled topologies relating five taxa.  We expected the
dominant topology across sampled genealogies to be concordant with that of the
species tree.  Indeed, two thirds of the local gene tree topologies (see section
\textbf{Materials and Methods}) along the alignment blocks were the same as the
species tree, which allowed us to be more confident of the species tree inferred
using the alignment blocks (Table S\ref{tab:Gene-tree-topologies}).  Other
frequently sampled topologies were expected to be potentially informative about
common recombination events.  The second most frequent topology (9.3\%) was one
where SDD was the outgroup, and SPY and SDE were sister clades. Therefore, SPY
and SDE appeared be more closely related with each other than each to SDD's
genome in about tenth of the core genome. The other topologies had less than 4\%
of proportions.

@@

The heat map of Figure \ref{fig:Heatmap-of-recombination} (see Tables
S\ref{tab:heatmap} and S\ref{tab:obsheatmap} for the heat map values and the
observed number of recombinant edges) shows the logarithm base 2 ratios of the
\textit{a posteriori} number of recombinant edges for a pair of donor and
recipient species tree branches and their corresponding expected values given
the species tree (Figure S\ref{fig:tree5}) and the three population parameters
(Table S\ref{tab:three}).  Because recombinant edges from any species tree
branch to the root branch are impossible (see Figure S\ref{fig:clonalorigin}),
the column for the root as recipient was colored green, indicating invalid
recombinant edges in Figure \ref{fig:Heatmap-of-recombination}.  
Similarly, all of the cells for which a recipient branch was
older than the donor, and the donor and recipient branches did not share any contemporary time were
colored green. The diagonal cells were colored gray, indicating no significant
number of recombinant edges, because the observed and expected numbers of
recombinant edges were similar. As recombinant edges that start from and end in
the same species tree branch would not change the species tree topology, which
might not
change the likelihood much, the cells would be colored gray.  Because SDD is a
strictly veterinary pathogenic bacteria, we did not expect that SDD received
much gene transfer from either SPY1 or SPY2: the logarithm base 2 ratio from SPY1 to
SDD was -2.20, and that from SPY2 to SDD was -1.97.  On the contrary, the
recombing gene transfer from SDE1 to SDD was 0.756, and that from SDE2 to SDD
was 0.698. This relatively higher recombining gene transfer from two SDE
individuals to SDD compared with that from two SPY individuals to SDD reflected
the fact
that the SDD was more closely related to SDE than to SPY in most of the genomes,
and the genomic organization between SDD and SDE1/SDE2 were much similar
(Figure S\ref{fig:mauve}).  However, the gene transfer from SDD to either SDE1
or SDE2 was less frequent than the case of reverse direction; SDD effectively
served as a recipient rather than a donor. It was not unclear why SDD received
more genes from SDE1 or SDE2 than did SDE1 or SDE2 from SDD.

Because it was expected that the recombining gene transfer within a clade would
be more frequent than that between clades, not surprisingly the recombining gene
transfer logarithm base 2 ratios between SDE1 and SDE2 (1.17 and 1.20) were about
twice as large as those from SDE1 and SDE2 to SPY1 and SPY2 (0.67, 0.66, 0.60,
and 0.39).  The intensities of recombining gene transfer from external branches
of SDE1 and SDE2 to external branches of SPY1 and SPY2 was similar to those from
SDE1 or SDE2 to SDD (0.76 and 0.70).  Therefore, either SPY1 or SPY2, and SDD as
recipient appeared to be no different from the view point of SDE1 or SDE2 as
donor since the similar values of recombining gene transfer intensity from SDE1
or SDE2 to one of the other three external branches. Surprisingly, the recombining
gene transfer logarithm base 2 ratios between SPY1 and SPY2 (1.06 and 1.07) was
comparable with, yet even less than those from SPY1 and SPY2 to SDE1 and SDE2
(1.25, 1.25, 1.17, and 1.29).  The recombining gene transfer from a clade SPY1 
or SPY2 to a different clade SDE1 or SDE2 seemed to be effectively no different
from that within the clade of SPY1 and SPY2.  The two external branches of SDE
received genes from either SPY1 or SPY2 via recombining gene transfer as if SDE1
and SDE2 were in the same clade as SPY1 and SPY2.  This was more surprising
because the bacterial recombination between SPY and SDE seemed to happen more
often than that between SDE and SDD even though the genome configurations of SDE
and SDD were more similar than each to SPY.  However, this result could be
concordant with the fact that SPY and SDE shared the same ecological niche
despite their dissimilar genomic organization.  The gene transfer logarithm base 2
ratio from SPY1 and SPY2 to SDE branch was also larger than any type of
recombining gene transfer: 1.88 and 1.94, respectively.  Additionally we found
58 genes for which a recombinant edge of certain types were sampled with
probability at least 0.9 by focusing on specific types of recombinant edges
(Table S\ref{tab:genes-transfer}).  Many of the genes in the list were
transferred from SPY to SDE rather than the reverse direction overall.  All
things considered, this epitomized that the recombining gene transfer was biased
toward the net direction from SPY to SDE.

We tried in vain to quantify differential horizontal gene transfer between SPY
and SDE with the parsimony-based approach (Figure \ref{fig:hgt-heatmap}).  Among
the clade-crossing transfers, we did not see any significant enrichment for a
specific direction, as opposed to the result for recombining transfers from the
model-based analysis.  SDE1 and SDE2 shared the most events (26 and 34) compared
to other sample pairs.  SDD appeared to acquire many genes from the ancestral
SDE branch.  Transfer events between the SPY and SDE clades were inferred in
smaller numbers (<6) compared to those within each clade. We were puzzled by
this finding because virulence genes were found to be associated with horizontal
gene transfer, and yet we could not show net directionality of horizontal gene
transfer between SPY and SDE.  However, we inferred relatively more recombining
gene transfer events from the SPY branch to SDE branch even in the
parsimony-based approach (Figure S\ref{fig:mowgli-recomb-heatmap}).

\subsection{Reduction of genes and divergence of SPY and SDE species}

The parsimony-based approach estimated 304 duplications, 944 recombining gene
transfers, and 294 horizontal gene transfers, comprising 23.7\% of total
transfer events (Figure \ref{fig:Gene-duplication-loss}).  Among the 2314 gene
families the largest family consisted of 59 genes, while nearly half of gene
families (1066) had exactly one gene per species, or six genes total (Figure
S\ref{fig:famsizes}).  Numbers of genes at the five ancestors were inferred to
be smaller than those at the extant species.  SDE lineages from the root
appeared to have increased in the number of genes while the common ancestor of
SPY1 and SPY2 decreased in the number of genes.

The four external lineages of SDE1, SDE2, SDD, and SPY1 had almost equal numbers
of gene gain and loss events while the four lineages of SP, SD, SPY2, and SEE
had larger numbers of gene gain events than those of gene loss events. In the
two lineages of SP and SEE we might have inferred gene gain and loss events less
confidently than events in other lineages because of lack of outgroup lineages
to compare with the two lineages. The number of gene gain events in SPY2 lineage
seemed to be large compared with those of gene gain events in the other
lineages. On the contrary, two internal branches of SPY and SDE had smaller
numbers of gene gain events than those of gene loss events.  This might indicate
that SPY and SDE lineages had reduced number of genes to specialize in their
specific niches after each split from their corresponding direct ancestors.

Duplications in extant lineages were more frequent than those in the other
lineages.  Although the number of duplication events along SPY2 branch seemed
smaller than those of the other extant lineages, duplication events appeared to
be most often species-specific.  Within clades we found that duplications made
up only a small fraction of the events  (10.2\%), while horizontal transfers and
losses were significantly more frequent (41.7\% and 48.0\%, respectively).  The
extant species of SDD and two SDE's were the most expanded, both in terms of
duplications and in-bound transfers. This was apparently related with the larger
genome sizes of the two species than that of SPY.  
The numbers of recombining gene transfer events tended to be larger than those
of horizontal gene transfer events except SD lineage.  Numbers of recombining
gene transfer in the extant lineages might have been larger than those in the
internal branches because the number of recombining gene transfer events must
had been affected by the availability of more closely related individual
genomes.

@@

About 10\% of the genes comprised of virulence facters, about half of which were
found in the core genome of the five genome alignment \citep{Suzuki2011}.  Using
customized UCSC genome browser tracks available at
\url{http://strep-genome.bscb.cornell.edu} we located virulence genes with
notable posterior recombination probability.  These genes were by no means a
comprehensive list of virulence factors that might have been transferred via
homologous recombination.  We wished to showcase virulence genes using the
result of our strudy of gene transfer.

Stressful host conditions for pathogenic bacteria can include heat shock,
starvation, and iron depletion, etc.  Iron uptake of bacteria was known to be
essential for survival of bacteria under the hostile host condition. We found
that parts of putative calcium transporter (MgtA) in SDE1 could have been
transferred from SPY1 (SpyM3\_0440).  An extracellular protein, putative
dipeptidase (SpyM3\_0465) that could be important for bacteria to scavenge
dipeptides from their host, was inferred to be transferred from SPY to SDE: the
whole gene was transferred, which was a case for the finding that virulence
genes tended to be transferred as a whole unit \citep{Chan2009}.  Another
example of whole gene transfer was superoxide dismutase that was reported to be
associated with virulence of \textit{Streptococcus agalactiae}
\citep{Poyart2001}: the whole gene including its 5' UTR, (SpyM3\_1071), was
affected by recombining gene transfer from SPY to SDE.  Some of the genes were
only affected in their parts.  The middle parts of an adhesins gene, putative
internalin A precursor (SpyM3\_1035), was found to be affected by recombining
gene transfer from SPY to SDE.  The middle part of a chaperonin GroEL, putative
heat shock protein (SpyM3\_1765), was inferred to transfer from SPY to SDE1:
increase of GroEL expression was known to allow for the bacteria to survive the
hostile host environment of acid shock.  Parts of the gene (SpyM3\_1749),
pyruvate formate lyase that was recently reported to have virulence
\citep{Yesilkaya2009}, seemed to be affected by recombining gene transfer.
Under starvation condition bacteria must use more effective energy metabolism: a
putative glucose kinase, (SpyM3\_1180), were inferred to be transferred from SPY
to SDE to adapt SDE's energy metabolism to survive such harsh condition.  We
also found that an MLST locus, called DNA mismatch repair protein (MutS),
appeared to be affected by recombining gene transfer.

\subsection{Association of virulence genes and horizontal gene transfer}

@@

Roughly, half of the virulence genes were in the pan genome as mentioned above.
We did not attempt to comprehensively list virulence genes that were inferred to
be transferred horizontally or non-recombining fashion. We rather sort virulence
genes by the number of horizontal gene transfer events to list the first several
genes with high ranks.  Streptococcal C5a peptidase (\textit{scpB}, SDEG\_0933)
was inferred to have six horizontal gene transfer.  Putative transposase were
found to have a relatively large number of horizontal gene transfer events
(SpyM3\_1016, SpyM3\_1053, SpyM3\_1054). Other genes with relatively large
number of horizontal gene transfer events included putative response regulator
(SpyM3\_0174), putative maltose operon transcriptional repressor (SpyM3\_0982),
putative histidine kinase (SpyM3\_0171), and putative cell wall hydrolase
(SpyM3\_0731, SpyM3\_0922, SpyM3\_1208).  Virulence genes encoding putative
nucleases with a secretion signal peptide were also found to have had
experienced horizontal gene transfer: genome-encoded streptodornase
(SDEG\_0541), and prophage-associated deoxyribonuclease (SDEG\_1103).  Protein
F2 like fibronectin-binding protein (SpyM3\_0104), an adhesin and invasin, was
inferred to horizontally transferred.

\subsection{Model-based approach or parsimony-based one for recombining gene
transfer study}

A full model-based approach would be desirable in analyzing data for testing
relevant hypotheses more systematically. Model complexity might not allow easy
access to the implementation of a model. Our study involved a model-based
approach and a parsimony-based one for understanding evolution in the core
genomes of several bacteria individuals. Because of inability of the model-based
method for studying the pan genomes we had to resort to the parsimony-based
method. However, the parsimony-based method was also applied to the core
genomes. We wished to compare the two methods in their results of the core
genome analyses.  Because of two independent analyses on the same data set, this
would allow us to be more confident of the results if the two analyses performed
similarly Generally, phylogenetic methods of detecting gene transfers often take
advantage of discrepancy between a reference species tree and gene trees. The
two approaches that we employed in this study were no exception. We found a few
similar results in both methods; firstly, functional categories that were
associated with recombining gene transfer were similar (Tables
S\ref{tab:functional-all} and S\ref{tab:go-events-recombining}).  Secondly,
either of them did not show association of virulence genes with recombining gene
transfer.  To more directly compare the methods we used the recombination
intensity and the number of recombining gene transfer events for each gene.  We
found a weak positive correlation of 0.39 with a very small significant P-value
of $2.2\times10^{-16}$ for testing a null hypothesis of the correlation being
zero.  (Figure S\ref{fig:cmpcomowgli}).  Interestingly, the correlation (0.03
with P-value of 0.18) disappeared in a comparison between recombination
intensity and number of horizontal gene transfer events (Figure
S\ref{fig:cmpcomowglihgt}).

\subsection{UCSC genome browser tracks for bacterial gene transfer study}

Often computational and statistical analyses on genomes could benefit other
researchers when those resources were provided: e.g., UCSC Genome Browser, and
Ensembl Genome Browser. We wished to provide research community especially in
SPY and SDE bacteria with some of the results from our analysis of gene
transfer. We used the UCSC genome browser for the purpose.  The resource shown
in Figure S\ref{fig:ucsc} for gene transfer is available at
\url{http://strep-genome.bscb.cornell.edu}.  For each of the five individual
genomes as a reference genome several annotation tracks were displayed:
alignment blocks, recombination rate per site for each block, local gene tree
topologies, recombination intensity, locations of virulence genes, posterior
probability of recombining gene transfer. The genome track of probability of
recombining gene transfer allowed us to easily browse the bacterial genome to
locate regions that we might be interested in. 
% To Matt: what could we show in the pan-genome? or from Mowgli's analysis?

@@

The process of gene transfer may or may not move whole genes from one species to
another \citep[e.g.,][]{Chan2009}. We observed that units of recombining gene
transfer could be multiple genes (Table S\ref{tab:genes-transfer}). In some
cases, parts of a gene were inferred to have been transferred, for instance,
transfer from SPY1 to SDE1 was inferred for 29\% of the aspartyl-tRNA synthetase
gene. Recombinant tracts often were shown to span two genes.  In other cases,
the inferred transfer units spaned a set of as much as eight adjacent genes:
loci from SpyM3\_1518 to SpyM3\_1525.

\subsection{Transporters, transferases, aminoacyl tRNA synthetases, and
transposases}
